A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-18 podc_paper/algorithm.tex
moved algorithm description into separate file algorithm.tex

%%
%% This is file generated with the docstrip utility from the source file: 
%%
%% samples.dtx  (with options: `acmsmall-conf')
%% With some modifications for PODC 2023.
%% For the copyright see the source file.
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,nonacm,anonymous]{acmart}

\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{relsize}
\usepackage{enumitem}
\usepackage{bold-extra}
\usepackage{showkeys}
\renewcommand*\contentsname{Table of Contents}

\usepackage{multicol}
\setlength\columnsep{24pt}

\algnewcommand\algorithmicforeach{\bf{for each}}
\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

\algdef{S}[FUNCTION]{Function}
   [3]{{\tt{\sl{#1}}} {\tt{#2}}\ifthenelse{\equal{#3}{}}{}{\tt{(#3)}}}

\algdef{E}[FUNCTION]{EndFunction}
   [1]{\algorithmicend\ \tt{{#1}}}

\algrenewcommand\Call[2]{\tt{{#1}\ifthenelse{\equal{#2}{}}{}{(#2)}}}

\newcommand\keywordfont{\sffamily\bfseries}
\algrenewcommand\algorithmicend{{\keywordfont end}}
\algrenewcommand\algorithmicfor{{\keywordfont for}}
\algrenewcommand\algorithmicforeach{{\keywordfont for each}}
\algrenewcommand\algorithmicdo{{\keywordfont do}}
\algrenewcommand\algorithmicuntil{{\keywordfont until}}
\algrenewcommand\algorithmicfunction{{\keywordfont function}}
\algrenewcommand\algorithmicif{{\keywordfont if}}
\algrenewcommand\algorithmicthen{{\keywordfont then}}
\algrenewcommand\algorithmicelse{{\keywordfont else}}
\algrenewcommand\algorithmicreturn{{\keywordfont return}}

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

\newcommand{\sub}[1]{\textsubscript{#1}}
\renewcommand{\tt}[1]{\texttt{#1}}
\renewcommand{\sl}[1]{\textsl{#1}}
\renewcommand{\it}[1]{\textit{#1}}
\renewcommand{\sc}[1]{\textsc{#1}}
\renewcommand{\bf}[1]{\textbf{#1}}
\newcommand{\nf}[1]{{\normalfont{\texttt{#1}}}}
\newcommand{\cmt}[1]{\Comment{#1}}
\newcommand{\head}{head}
\newcommand{\size}{size }

\newcommand{\here}[1]{\bf{[[#1]]}}

\usepackage{amsmath,amsthm}
%\newtheorem{theorem}{Theorem}
%\newtheorem{lemma}[theorem]{Lemma}
%\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{observation}{Observation}
%\theoremstyle{definition}
%\newtheorem{definition}[theorem]{Definition}
\newtheorem{invariant}{Invariant}
%\newtheorem{proposition}[theorem]{Proposition}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none}  %FIX FOR CAMERA READY
%\copyrightyear{2023}
%\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}
\fancyfoot{} % REMOVE THIS LINE FOR CAMERA READY

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[PODC 2023]{ACM Symposium on Principles of Distributed Computing}{June 19--23,
  2023}{Orlando, FL}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
%\acmPrice{15.00}
%\acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

\title{A Wait-free Queue with Poly-logarithmic Step Complexity}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Eric Ruppert}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}
\email{ruppert@cse.yorku.ca}
\orcid{1234-5678-9012}

\author{Hossein Naderibeni}
\email{hnaderi268@gmail.com}
\affiliation{%
  \institution{York University}
  \streetaddress{P.O. Box 1212}
  \city{Toronto}
  \state{Ontario}
  \country{Canada}
  \postcode{43017-6221}
}

\settopmatter{printacmref=false}  % REMOVE THIS LINE FOR CAMERA READY

\begin{abstract}
We introduce a novel linearizable wait-free queue implementation using single-word 
CAS instructions.
Previous lock-free queues all take $\Omega(p)$ steps per operation in the worst case, 
where $p$ is the number of processes that can access the queue.
We achieve $O(\log^2 p +\log q)$ steps per operation, where $q$ is the size of the queue.
\here{mention space usage; update time bound if space-efficient version takes more time per op}
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{concurrent data structures, wait-free queues}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\input{introduction}
\input{related}

@@

In our model there are $p$ processes doing \nf{Enqueue} and \nf{Dequeue} operations on a queue concurrently. We design a linearizable wait-free queue with $O(\log^2 p +\log q)$ steps per operation, where $q$ is the number of elements in the queue at the time of linearization. We avoid the $\Omega(p)$ worst-case step complexity of existing shared queues based on linked lists or arrays, which suffer from the CAS Retry Problem.

There is a shared binary tree among the processes to agree on one total ordering of the operations invoked by processes. The tree is called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each process has a leaf in which the operations invoked by the process are stored in order. When a process wishes to do an operation it appends the operation to its leaf and tries to propagate its new operation up to the tree's root. Each node of the tree keeps an ordering of operations propagated up to it. All processes agree on the sequence of operations in the root and this ordering is used as the linearization ordering. 
\begin{figure}[hbtp]
\begin{center}
\Tree [ [ [ $P_1$ $P_2$ ] [ $P_3$ $P_4$ ] ]
          [ [ $P_5$ $P_6$ ] [ $P_7$ $P_8$ ] ] ]
\end{center}
\caption[Tournament tree.]{\label{fig::blocktree}Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node there is an ordering of operations stored. Each process tries to propagate its operations up to the root, which stores a total ordering of all operations.}  
\end{figure}

@@

\subsection{Pseudocode} \label{algQ}
We present our algorithm in pseudocode. page~22 contains the description of the fields in the tree nodes and the blocks. The value of any uninitialized field is \nf{null}. page~23 contains major routines and the rest of this section consists of the auxiliary routines. The abbreviations below are used in the pseudocode and the proof of correctness.
\begin{itemize}
 \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \tt{\{enq, deq\}})
%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
\end{itemize}

\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \textsf{A binary tree of \tt{Node}s with one \tt{leaf} for each process. \tt{root} is the root \nf{node}.}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf:} \sf{ process's leaf in the tree.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: Initialized  when creating the tree.}
\item \tt{\sl{Block[]} blocks} \textsf{: Initially \tt{blocks[0]} contains an empty block with all fields equal to 0.}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Block}}

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{:~~indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{: \#enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{: \#dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{: \#enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{: \#dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{: Each block in a leaf represents a single operation. If the operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{: \# enqueue, dequeue operations in this block and its previous blocks in the leaf}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} \size}
  \textsf{: size of the queue after performing all operations in this block and its previous blocks in the root}
\end{itemize}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Queue}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{0}

\Function{void}{Enqueue}{\sl{Object} e} \cmt{Creates a \tt{block} with element \tt{e} and adds it to the tree.}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}}
\State \tt{newBlock.element= e}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}+1}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}}
\State \tt{leaf.}\Call{Append}{newBlock}
\EndFunction{Enqueue}

\Statex

\Statex $\triangleright$ Creates a block with \nf{null} value element, appends it to the tree and returns its response.
\Function{Object}{Dequeue()}{} 
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} 
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.\head].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.\head].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b, i>=} \Call{IndexDequeue}{leaf.\head, 1}
\State \tt{output=} \Call{FindResponse}{b, i} 
\label{deqRest}
\State \Return{\tt{output}}
\EndFunction{Dequeue}

\Statex

\Statex $\triangleright$ Returns the response to $D_i(root,b)$, the \nf{i}th \nf{Dequeue} in \nf{root.blocks[b]}.
\Function{element}{FindResponse}{\sl{int} b, \sl{int} i}
\If{\tt{ root.blocks[b-1].\size}\tt{ + root.blocks[b].num\sub{enq} - i $<$ 0}} \label{checkEmpty}\cmt{Check if the queue is empty.}
\State \Return \tt{null} \label{returnNull}
\Else \cmt{The response is $E_e(root)$, the \nf{e}th \nf{Enqueue} in the root.}
\State \tt{e= i + (root.blocks[b-1].sum\sub{enq}-root.blocks[b-1].size)} \label{computeE}
\State \Return \tt{root.GetEnqueue(root.\Call{DoublingSearch}{e, b})}\label{findAnswer}
\EndIf
\EndFunction{FindResponse}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{25}

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with \tt{sum\sub{enq}} greater than or equal to \tt{x}}
\Statex $\triangleright$ \textmd{Does a binary search for~the value \tt{x} of \tt{sum\sub{enq}} field and returns the index of the leftmost block in $\triangleright$~\tt{blocks[start..end]} whose \tt{sum\sub{enq}} is $\geq$ \tt{x}}.
\Function{int}{BinarySearch}{\sl{int} x, \sl{int} start, \sl{int} end}
% \State \Return \tt{min\{j: blocks[j].sum\sub{enq}$\geq$x\}}
\While{\nf{start<end}}
\State \tt{\sl{int}} \tt{mid= floor((start+end)/2)}
\If{\nf{blocks[mid].sum\sub{enq}<x}}
\State \nf{start= mid+1}
\Else
\State \nf{end= mid}
\EndIf
\EndWhile
\State\Return \nf{start}
\EndFunction{BinarySearch}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Root}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{36}
\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{e}}}
\Statex $\triangleright$ \textmd{Returns \tt{<b,i>} such that $E_\nf{e}(\nf{root})=E_\nf{i}(\nf{root},\nf{b})$, i.e., the \nf{e}th \nf{Enqueue} in the \nf{root} is the \nf{i}th \nf{Enqueue} within $\triangleright$~the \nf{b}th block in the \nf{root}.}

\Function{<int, int>}{DoublingSearch}{\sl{int} e, \sl{int} end}
\State \tt{start= end-1} \label{dsearchStart}
\While{\tt{root.blocks[start].sum\sub{enq}}$>=$\tt{e}}
\State \tt{start= max(start-(end-start), 0)} \label{doubling}
\EndWhile \label{dsearchEnd}
\State \tt{b= root.BinarySearch(e, start, end)} \label{dsearchBinarySearch}
\State \tt{i= e- root.blocks[b-1].sum\sub{enq}} \label{DSearchComputei}
\State\Return \tt{<b,i>}
\EndFunction{DoublingSearch}
\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Leaf}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{45}

\Function{void}{Append}{\sl{block} B} \cmt{Only called by the owner of the leaf.}
\State \tt{blocks[\head]= B} \label{appendLeaf}
\State \tt{\head= \head+1} \label{appendEnd} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{50}

\Statex $\triangleright$ \textmd{\nf{$n$.Propagate} propagates operations  in \nf{this}.children up to \nf{this} when it terminates.}
\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}} \label{firstRefresh}
\State \Call{Refresh()}{} \label{secondRefresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

\Statex

\Statex $\triangleright$ \textmd{Creates a block containing new operations of \nf{this.}children, and then tries to append it to \nf{this}.}
\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{startHelpChild1}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\If{\nf{dir.blocks[h\sub{dir}]!=null}} \label{ifHeadnotNull}
\State{\tt{dir.\Call{Advance}{h\sub{dir}}}} \label{helpAdvance}
\EndIf
\EndFor \label{endHelpChild1}
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} 
\EndIf
\State{\tt{result= blocks[h].CAS(null, new)}} \label{cas}
\State{\tt{this.\Call{Advance}{h}}} \label{advance}
\State \Return{ \tt{result}}

\EndFunction{Refresh}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{73}

\Function{void}{Advance}{\sl{int} h} \cmt{Sets \nf{blocks[h].super} and increments \nf{head} from \nf{h} to \nf{h+1}.}
\State \tt{h\sub{p}= parent.\head} \label{readParentHead}
\State \tt{blocks[h].super.CAS(null, h\sub{p})} \label{setSuper1}
\State \tt{head.CAS(h, h+1)} \label{incrementHead}
\EndFunction{Advance}

\Statex

\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
\EndIf

\State \Return \tt{new}
\EndFunction{CreateBlock}

\end{algorithmic}
\end{algorithm}

\begin{algorithm}
\caption{\tt{\sl{Node}}}
\begin{algorithmic}[1]
\setcounter{ALG@line}{94}

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
\EndIf
\EndFunction{GetEnqueue}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_\tt{i}(\tt{this}, \tt{b})=D_\tt{y}(\tt{root},\tt{x})$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n ? left: right)} 
\State \tt{superblockIndex= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ? \label{computeSuper}
\Statex \hspace{11.3em} blocks[b].super: blocks[b].super+1}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-left}} \label{considerPreviousLeft}
\Else \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[superblockIndex-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= parent.blocks[superblockIndex].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{superblockIndex, i}
\EndIf
\EndFunction{IndexDequeue}

\end{algorithmic}
\end{algorithm}

\section{Proof of Correctness}

We adopt linearizability as our definition of correctness. In our case, where we create the linearization ordering in the root, we need to prove (1) the ordering is legal, i.e, for every execution on our queue if operation $op_1$ terminates before operation $op_2$ then $op_1$ is linearized before operation $op_2$ and (2) if we do operations sequentially in their linearization order, operations get the same results as in our queue. The proof is structured like this. First, we define and prove some facts about blocks and the node's \nf{head} field. Then, we introduce  the linearization ordering formally. Next, we prove double \nf{Refresh} on a node is enough to propagate its children's new operations up to the node, which is used to prove (1). After this, we prove some claims about the size and operations of each block, which we use to prove the correctness of \nf{DoublingSearch()}, \nf{GetEnqueue()} and \nf{IndexDequeue()}. Finally, we prove the correctness of the way we compute the response of a dequeue, which establishes (2).

\subsection{Basic Properties}
In this subsection, we talk about some properties of blocks and fields of the tree nodes.

A block is an object storing some statistics, as described in Algorithm Queue. A block in a node implicitly represents a set of operations.
\begin{definition}[Ordering of a block in a node]
Let $B$ be \nf{$n$.blocks[$b$]} and $B^\prime$ be \nf{$n$.blocks[$j$]}. We call $i$ the \emph{index} of block $B$. Block $B$ is \it{before} block $B^\prime$ in node $n$ if and only if $i<j$.
\end{definition}

Next, we show that the value of \nf{head} in a node can only be increased. By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
\begin{observation} \label{nonDecreasingHead}
For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
\end{proof}

\begin{lemma} \label{lem::headInc}
Let $R$ be an instance of \nf{Refresh} on a node $n$ that creates a non-empty block \nf{new} ($\nf{new.sum}\neq 0$). After $R$ terminates, \nf{$n$.head} is greater than the value read in Line \ref{readHead} of $R$.
\end{lemma}
\begin{proof}
If the \nf{CAS} in Line \ref{incrementHead} is successful, then the claim holds. Otherwise, $n.\nf{head}$ has changed from the value read in Line \ref{readHead}. By Observation \ref{nonDecreasingHead} this means another process has incremented $n.\nf{head}$.
\end{proof}

Now we show \nf{$n$.blocks[$n$.head]} is either the last block written into node $n$ or the first empty block in~$n$.

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of $n.\nf{head}$ is $h$ then $\nf{$n$.blocks[$i$]}=\nf{null}$ for ${i>h}$ and $\nf{$n$.blocks[$i$]}\neq\nf{null}$ for ${0\leq i<h}$.
\end{invariant}
\begin{proof}
Initially the invariant is true since $n.\nf{head}=1$, $n.\nf{blocks[0]}\neq\nf{null}$  and $n.\nf{blocks[$x$]}=\nf{null}$ for every $x>0$. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}. We show that if the invariant holds before such a change, then it still holds after the change.

In the algorithm, $n.\nf{blocks}$ is modified only on Line \ref{cas}, which updates $n.\nf{blocks[$h$]}$ where $h$ is the value read from $n.\nf{head}$ in Line \ref{readHead}. Since the \nf{CAS} in Line \ref{cas} is successful, it means $n.\nf{head}$ has not changed from $h$ before doing the \nf{CAS}: if $n$\nf{.head} had changed before the \nf{CAS} then it would be greater than $h$ by Observation~\ref{nonDecreasingHead} and hence $n$\nf{.blocks[}$h$\nf{]}$\neq$\nf{null} and by the induction hypothesis, so the \nf{CAS} would fail. Writing into \nf{$n$.blocks[$h$]} when $h=n$\nf{.head} preserves the invariant, since the claim does not talk about the content of \nf{$n$.blocks[$n$.head]}.

The value of \nf{$n$.head} is modified only in Line \ref{incrementHead} of \nf{Advance}. If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. \nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}. For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}. For Line \ref{advance}, Line \ref{cas} was finished  before doing Line \ref{advance}. Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}. 
\end{proof}

We define the subblocks of a block recursively.

\begin{definition}[Subblock]\label{def::subblock}
A block is a \emph{direct subblock} of the $i$th block in node $n$ if it is\\ in $$\texttt{$n$.left.blocks[$n$.blocks[$i-1$].end\textsubscript{left}+1$\cdots n$.blocks[$i$].end\textsubscript{left}]}$$ or in $$\texttt{$n$.right.blocks[$n$.blocks[$i-1$].end\textsubscript{right}+1$\cdots n$.blocks[$i$].end\textsubscript{right}]}.$$
Block $B$ is a \emph{subblock} of block $C$ if $B$ is a direct subblock of $C$ or a subblock of a direct subblock of~$C$.  We say block $B$ has been \it{propagated} to node $n$ if $B$ is in $n.\nf{blocks}$ or is a subblock of a block in $n\nf{.blocks}$.
\end{definition}

The following lemma is used to prove that subblocks of two blocks in a node are disjoint.
\begin{lemma} \label{lem::headProgress}
 If $n\nf{.blocks[}b\nf{]}\neq\nf{null}$$(b>0)$ then $\nf{$n$.blocks[$i$].end\sub{left}} \geq \nf{$n$.blocks[$i-1$].end\sub{left}}$ and $\nf{$n$.blocks[$i$].end\sub{right}} \geq \nf{$n$.blocks[$i-1$].end\sub{right}}$.
\end{lemma}
\begin{proof}
Consider the block $B$ written into \nf{n.blocks[$b$]} by \nf{CAS} at Line \ref{cas}. Block $B$ is created by the \nf{CreateBlock($b$)} called at Line \ref{invokeCreateBlock}. Prior to this call to \nf{CreateBlock($b$)}, $\nf{n.head}=b$ at Line \ref{readHead}, so \nf{n.blocks[$b-1$]} is already a non-null value $B^\prime$ by Invariant \ref{lem::headPosition}. Thus, the \nf{CreateBlock($b-1$)} that created  $B^\prime$ has terminated before the invocation of \nf{CreateBlock($b$)} that created $B$. The value written into \nf{$B$.end\sub{left}} at Line \ref{lastLine} of \nf{CreateBlock($b$)}  was one less than the value of $n$\nf{.left.head} read at Line \ref{lastLine} of \nf{CreateBlock($b$)}. Similarly, the value in \nf{n.blocks[$b-1$].end\sub{left}} was one less than the value read from $\nf{n.left.head}$ during the call to \nf{CreateBlock($b-1$)}. By Observation \ref{nonDecreasingHead}, $n\nf{.left.head}$ is non-decreasing, so $B^\prime\nf{.end\sub{left}} \leq \nf{$B$.end\sub{left}}$. The proof for \nf{end\sub{right}} is similar.
\end{proof}

\begin{lemma} \label{lem::subblocksDistinct}
The sets of subblocks of any two blocks in a node are disjoint.  
\end{lemma}
\begin{proof}
We are going to prove the lemma by contradiction. Consider the lowest \nf{node} $n$ in the tree that violates the claim. Then, subblocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ overlap for some $i<j$. Since $n$ is the lowest node in the tree violating the claim, direct subblocks of blocks of $n\nf{.blocks[}i\nf{]}$ and $n\nf{.blocks[}j\nf{]}$ have to overlap. Without loss of generality, assume left child subblocks of $n\nf{.blocks[}i\nf{]}$ overlap with the left child subblocks of $n\nf{.blocks[}j\nf{]}$. By Lemma \ref{lem::headProgress} we have $ \nf{$n$.blocks[$i$].end\sub{left}} \leq \nf{$n$.blocks[$j-1$].end\sub{left}}$, so the range $[\nf{$n$.blocks[$i-1$].end\sub{left}}+1 \cdots \nf{$n$.blocks[$i$].end\sub{left}}]$ cannot have overlap with the range $[\nf{$n$.blocks[$j-1$].end\sub{left}}+1 \cdots \\ \nf{$n$.blocks[$j$].end\sub{left}}]$. Therefore, direct subblocks of \nf{$n$.blocks[$i$]} and \nf{$n$.blocks[$j$]} cannot overlap, which is in contradiction with the assumption.
\end{proof}

>>>>

% !TEX root =  podc-submission.tex

@@

We use a static {\it tournament tree}, which is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
assigned to each process.

There is a shared binary tree among the processes to agree on one
total ordering of the operations invoked by processes. The tree is
called a \it{tournament tree} (see Figure~\ref{fig::blocktree}). Each
process has a leaf in which the operations invoked by the process are
stored in order. When a process wishes to do an operation it appends
the operation to its leaf and tries to propagate its new operation up
to the tree's root. Each node of the tree keeps an ordering of
operations propagated up to it. All processes agree on the sequence of
operations in the root and this ordering is used as the linearization
ordering.

Each of the processes $P_1,P_2,...,P_p$ has a leaf and in each node
there is an ordering of operations stored. Each process tries to
propagate its operations up to the root, which stores a total ordering
of all operations.

@@
